---
title: "Lab 06 — Hypothesis Testing from Claims"
author: "EAES 480 — Modern Statistics in Earth & Environmental Science"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  html_document:
    theme: cosmo
    highlight: tango
    toc: true
    toc_depth: 3
    toc_float: true
    code_folding: hide
    df_print: paged
editor_options:
  chunk_output_type: console
---

```{r setup, echo=FALSE, message=FALSE, warning=FALSE}
library(tidyverse)
library(lubridate)
library(janitor)

knitr::opts_chunk$set(
  warning = FALSE,
  message = FALSE
)
```

# Overview

This lab introduces **hypothesis testing** as an extension of sampling, bootstrap distributions, and uncertainty quantification.

The core idea is simple:

> A **claim** about the population becomes a **null hypothesis**.

You will use the **US-AMS simplified AmeriFlux dataset** again and test claims about population means using:

- one sample,
- a bootstrap estimate of uncertainty,
- a z-statistic,
- and a p-value.

This lab emphasizes **reasoning from claims to tests**, not just mechanics.

---

# Learning goals

By the end of this lab, you should be able to:

- Turn a verbal claim into a null hypothesis
- Choose a variable and define the relevant population parameter
- Draw a sample and estimate a mean
- Use bootstrap resampling to estimate uncertainty in that mean
- Compute a z-statistic and p-value
- Interpret a hypothesis test in plain language
- Connect confidence intervals and hypothesis tests

---

# Data

## Load and inspect

```{r load_data, echo=TRUE, eval=FALSE}
# GOAL: Read the dataset and clean the names.

df <- read_csv("data/us-ams-simple.csv", na = c("-9999")) %>%
  clean_names() %>%
  filter(year_local == 2023)

glimpse(df)
```

## Derive time columns (if needed)

```{r derive_time, echo=TRUE, eval=FALSE}
# GOAL: Create date / month / day_night columns as in earlier labs.

df <- df %>%
  mutate(
    date = as.Date(floor(doy) - 1, origin = paste0(year_local, "-01-01")),
    month = month(date, label = TRUE, abbr = TRUE),
    day_night = if_else(daytime == 1, "Day", "Night")
  )
```

---

# Part I — From claim to hypothesis test

## 1) Choose a variable

Choose **one numeric variable** from the dataset.

Possible examples:
- `swc`
- `ta`
- `fc`
- `gpp`
- `le`

```{r choose_variable, echo=TRUE, eval=FALSE}
target_var <- "gpp"

# CHECK: does this variable exist?
target_var %in% names(df)
```

## 2) Describe the population parameter

Write the population parameter you care about:

> $\mu$ = the true 2023 population mean of `gpp`

Fill in the blank in words:

> In this lab, my population parameter is the true annual mean of **`gross primary production`** at the US-AMS site.

---

## 3) Make a claim

Write a claim about the population mean for your variable.

Examples:
- “The annual mean soil water content is 0.30.”
- “The annual mean air temperature is 10 °C.”
- “The annual mean latent heat flux is 40 W m$^{-2}$.”

*The annual mean gpp is 25 tons of carbon per hectre per year*

Now turn your claim into hypotheses:

- Null hypothesis: $H_0: \mu = \mu_0$
- Alternative hypothesis: $H_A: \mu \neq \mu_0$

Fill these in:

> My claim is: `The annual mean gpp is 25 tons C / ha / yr`

> $H_0$: 
        `$H_0: \mu = 25$`

> $H_A$: 
        `$H_A: \mu \neq 25$`

---

## 4) Draw one sample and compute the sample mean

```{r draw_sample, echo=TRUE, eval=FALSE}
# GOAL: Draw one random sample and compute its mean.

set.seed(560)

n_samp <- 231   # e.g., 200

samp <- df %>%
  filter(!is.na(.data[[target_var]])) %>%
  slice_sample(n = n_samp)

sample_mean <- mean(samp[[target_var]], na.rm = TRUE)

sample_mean
```

**Prompt (2–3 sentences):** How does your sample mean compare to your claim? Is it close, or already quite different?

> *The actual result is very different from my claim. My initial claim was that the mean annual gpp would be 25, while the actual value was ~6*

---

## 5) Bootstrap the sample mean

```{r bootstrap_means, echo=TRUE, eval=FALSE}
# GOAL: Bootstrap the sample mean to estimate uncertainty.

set.seed(5694)

B <- 3752   # e.g., 4000

boot_means <- replicate(
  B,
  mean(sample(samp[[target_var]], size = n_samp, replace = TRUE), na.rm = TRUE)
)

head(boot_means)
```

## 6) Estimate uncertainty

```{r bootstrap_summary, echo=TRUE, eval=FALSE}
# GOAL: Compute the bootstrap SE and 95% CI.

boot_se <- sd(boot_means)
boot_ci <- quantile(boot_means, probs = c(0.025, 0.975))

tibble(
  sample_mean = sample_mean,
  boot_se = boot_se,
  ci_lower = boot_ci[1],
  ci_upper = boot_ci[2]
)
```

**Prompt (2–3 sentences):** What does the bootstrap SE tell you? What does the 95% CI tell you?

> *The standard error tells us about the spread of the bootstrap. A higher SE means a higher spread, and a lower precision. The 95% confidence interval tells us that 95% of the time our sample will fall inbetween the bounds of the interval. A lower difference in upper and lower bounds means higher precision*

---

## 7) Convert the claim into a formal test

Define the null value from your claim.

```{r hypothesis_test, echo=TRUE, eval=FALSE}
# GOAL: Compute z and p-value for your claim.

mu0 <- 25   # numeric claim value

z_stat <- (sample_mean - mu0) / boot_se

p_value <- 2 * pnorm(-abs(z_stat))

tibble(
  sample_mean = sample_mean,
  mu0 = mu0,
  boot_se = boot_se,
  z_stat = z_stat,
  p_value = p_value
)
```

---

## 8) Interpret the result

Use your p-value and CI to answer:

- Does the 95% CI include the null value?
- Is the p-value small or large?
- Would you reject the null hypothesis at $\alpha = 0.05$?

Fill in:

> The null value `25` is not inside my 95% CI.

> My p-value is `7.98 * 10^(-185)`, which is small / large.

> Therefore, I would reject $H_0$ at the 0.05 level.

**Prompt (4–6 sentences):** Explain your decision in plain language. Do the data make the claim look plausible or surprising? Be careful: “fail to reject” does not mean “prove true.”

> *My decision to reject the null hypothesis comes from having such a small p-value. Our p-value tells us the likelihood of our suggested mean being the true mean of the population. A higher p-value means that we are more likely to have selected the actual mean value. Because my p-value is so small, the likelihood of my prediction being the actual mean gpp is excessively small as well. Due to this fact I have chosen to reject the "null hypothesis", or the idea that my predicted value is the true value.*

---

## 9) Visualize the hypothesis test

```{r plot_test, echo=TRUE, eval=FALSE}
# GOAL: Plot the bootstrap distribution, sample mean, CI, and null value.

ggplot(tibble(boot_mean = boot_means), aes(x = boot_mean)) +
  geom_histogram(bins = 50, alpha = 0.8) +
  geom_vline(xintercept = sample_mean, linetype = "dashed", linewidth = 1.1) +
  geom_vline(xintercept = boot_ci, color = "red", linewidth = 1.1) +
  geom_vline(xintercept = mu0, color = "blue", linewidth = 1.1) +
  theme_classic(base_size = 18) +
  labs(
    x = paste0("Bootstrap mean of ", target_var),
    y = "Count",
    title = "Bootstrap distribution and null claim",
    subtitle = "Dashed = sample mean; Red = 95% CI; Blue = null value"
  )
```

**Prompt (2–3 sentences):** Does the plot agree with your p-value-based conclusion? Explain.

> *Yes my plot agrees with my conclusion. My plot shows that my predicted value is far higher than the mean, and sits well outside of the 95% confidence interval. The likelihood of the true value being this high is essentially 0.*

---

## 10) Try a second claim (short extension)

Keep the same variable, but test a **different null value**.

```{r second_claim, echo=TRUE, eval=FALSE}
# GOAL: Test a second claim using the same sample and bootstrap SE.

mu0_alt <- 6

z_stat_alt <- (sample_mean - mu0_alt) / boot_se
p_value_alt <- 2 * pnorm(-abs(z_stat_alt))

tibble(
  mu0_alt = mu0_alt,
  z_stat_alt = z_stat_alt,
  p_value_alt = p_value_alt
)
```

**Prompt (3–4 sentences):** How did changing the null value change the conclusion? What does this tell you about the relationship between the null hypothesis and the observed estimate?

> *Changing the null value vastly improved my p-value. This tells me that the relationship between the null hypothesis and the observed estimate is dependant on the predicted null hypothesis. The observed estimate will not change unless more samples are taken, or samples are removed. Thus, the null hypothesis, which is variable and dependent on who is doing the prediction, is the determining factor for their relationship.*

---

# Part II — To be added later

We will add additional exercises here on:

- comparing means between groups,
- one- vs two-sided tests,
- and connections to t-tests / ANOVA.

> *Instructor will add this section later. Do not delete this header.*

---

# Submission

- Knit your `.Rmd` to HTML.
- Commit and push both the `.Rmd` and knitted `.html` to your GitHub repo.
- Submit the GitHub repository link on Canvas.
