---
title: "Lab 06 — Hypothesis Testing from Claims"
author: "EAES 480 — Modern Statistics in Earth & Environmental Science"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  html_document:
    theme: cosmo
    highlight: tango
    toc: true
    toc_depth: 3
    toc_float: true
    code_folding: hide
    df_print: paged
editor_options:
  chunk_output_type: console
---

```{r setup, echo=FALSE, message=FALSE, warning=FALSE}
library(tidyverse)
library(lubridate)
library(janitor)

knitr::opts_chunk$set(
  warning = FALSE,
  message = FALSE
)
```

# Overview

This lab introduces **hypothesis testing** as an extension of sampling, bootstrap distributions, and uncertainty quantification.

The core idea is simple:

> A **claim** about the population becomes a **null hypothesis**.

You will use the **US-AMS simplified AmeriFlux dataset** again and test claims about population means using:

- one sample,
- a bootstrap estimate of uncertainty,
- a z-statistic,
- and a p-value.

This lab emphasizes **reasoning from claims to tests**, not just mechanics.

---

# Learning goals

By the end of this lab, you should be able to:

- Turn a verbal claim into a null hypothesis
- Choose a variable and define the relevant population parameter
- Draw a sample and estimate a mean
- Use bootstrap resampling to estimate uncertainty in that mean
- Compute a z-statistic and p-value
- Interpret a hypothesis test in plain language
- Connect confidence intervals and hypothesis tests

---

# Data

## Load and inspect

```{r load_data, echo=TRUE, eval=FALSE}
# GOAL: Read the dataset and clean the names.

df <- read_csv("data/us-ams-simple.csv", na = c("-9999")) %>%
  clean_names() %>%
  filter(year_local == 2023)

glimpse(df)
```

## Derive time columns (if needed)

```{r derive_time, echo=TRUE, eval=FALSE}
# GOAL: Create date / month / day_night columns as in earlier labs.

df <- df %>%
  mutate(
    date = as.Date(floor(doy) - 1, origin = paste0(year_local, "-01-01")),
    month = month(date, label = TRUE, abbr = TRUE),
    day_night = if_else(daytime == 1, "Day", "Night")
  )
```

---

# Part I — From claim to hypothesis test

## 1) Choose a variable

Choose **one numeric variable** from the dataset.

Possible examples:
- `swc`
- `ta`
- `fc`
- `gpp`
- `le`

```{r choose_variable, echo=TRUE, eval=FALSE}
target_var <- "gpp"

# CHECK: does this variable exist?
target_var %in% names(df)
```

## 2) Describe the population parameter

Write the population parameter you care about:

> $\mu$ = the true 2023 population mean of `gpp`

Fill in the blank in words:

> In this lab, my population parameter is the true annual mean of **`gross primary production`** at the US-AMS site.

---

## 3) Make a claim

Write a claim about the population mean for your variable.

Examples:
- “The annual mean soil water content is 0.30.”
- “The annual mean air temperature is 10 °C.”
- “The annual mean latent heat flux is 40 W m$^{-2}$.”

*The annual mean gpp is 25 tons of carbon per hectre per year*

Now turn your claim into hypotheses:

- Null hypothesis: $H_0: \mu = \mu_0$
- Alternative hypothesis: $H_A: \mu \neq \mu_0$

Fill these in:

> My claim is: `The annual mean gpp is 25 tons C / ha / yr`

> $H_0$: 
        `$H_0: \mu = 25$`

> $H_A$: 
        `$H_A: \mu \neq 25$`

---

## 4) Draw one sample and compute the sample mean

```{r draw_sample, echo=TRUE, eval=FALSE}
# GOAL: Draw one random sample and compute its mean.

set.seed(560)

n_samp <- 231   # e.g., 200

samp <- df %>%
  filter(!is.na(.data[[target_var]])) %>%
  slice_sample(n = n_samp)

sample_mean <- mean(samp[[target_var]], na.rm = TRUE)

sample_mean
```

**Prompt (2–3 sentences):** How does your sample mean compare to your claim? Is it close, or already quite different?

> *The actual result is very different from my claim. My initial claim was that the mean annual gpp would be 25, while the actual value was ~6*

---

## 5) Bootstrap the sample mean

```{r bootstrap_means, echo=TRUE, eval=FALSE}
# GOAL: Bootstrap the sample mean to estimate uncertainty.

set.seed(5694)

B <- 3752   # e.g., 4000

boot_means <- replicate(
  B,
  mean(sample(samp[[target_var]], size = n_samp, replace = TRUE), na.rm = TRUE)
)

head(boot_means)
```

## 6) Estimate uncertainty

```{r bootstrap_summary, echo=TRUE, eval=FALSE}
# GOAL: Compute the bootstrap SE and 95% CI.

boot_se <- sd(boot_means)
boot_ci <- quantile(boot_means, probs = c(0.025, 0.975))

tibble(
  sample_mean = sample_mean,
  boot_se = boot_se,
  ci_lower = boot_ci[1],
  ci_upper = boot_ci[2]
)
```

**Prompt (2–3 sentences):** What does the bootstrap SE tell you? What does the 95% CI tell you?

> *The standard error tells us about the spread of the bootstrap. A higher SE means a higher spread, and a lower precision. The 95% confidence interval tells us that 95% of the time our sample will fall inbetween the bounds of the interval. A lower difference in upper and lower bounds means higher precision*

---

## 7) Convert the claim into a formal test

Define the null value from your claim.

```{r hypothesis_test, echo=TRUE, eval=FALSE}
# GOAL: Compute z and p-value for your claim.

mu0 <- 25   # numeric claim value

z_stat <- (sample_mean - mu0) / boot_se

p_value <- 2 * pnorm(-abs(z_stat))

tibble(
  sample_mean = sample_mean,
  mu0 = mu0,
  boot_se = boot_se,
  z_stat = z_stat,
  p_value = p_value
)
```

---

## 8) Interpret the result

Use your p-value and CI to answer:

- Does the 95% CI include the null value?
- Is the p-value small or large?
- Would you reject the null hypothesis at $\alpha = 0.05$?

Fill in:

> The null value `25` is not inside my 95% CI.

> My p-value is `7.98 * 10^(-185)`, which is small.

> Therefore, I would reject $H_0$ at the 0.05 level.

**Prompt (4–6 sentences):** Explain your decision in plain language. Do the data make the claim look plausible or surprising? Be careful: “fail to reject” does not mean “prove true.”

> *My decision to reject the null hypothesis comes from having such a small p-value. Our p-value tells us the likelihood of our suggested mean being the true mean of the population. A higher p-value means that we are more likely to have selected the actual mean value. Because my p-value is so small, the likelihood of my prediction being the actual mean gpp is excessively small as well. Due to this fact I have chosen to reject the "null hypothesis", or the idea that my predicted value is the true value.*

---

## 9) Visualize the hypothesis test

```{r plot_test, echo=TRUE, eval=FALSE}
# GOAL: Plot the bootstrap distribution, sample mean, CI, and null value.

ggplot(tibble(boot_mean = boot_means), aes(x = boot_mean)) +
  geom_histogram(bins = 50, alpha = 0.8) +
  geom_vline(xintercept = sample_mean, linetype = "dashed", linewidth = 1.1) +
  geom_vline(xintercept = boot_ci, color = "red", linewidth = 1.1) +
  geom_vline(xintercept = mu0, color = "blue", linewidth = 1.1) +
  theme_classic(base_size = 18) +
  labs(
    x = paste0("Bootstrap mean of ", target_var),
    y = "Count",
    title = "Bootstrap distribution and null claim",
    subtitle = "Dashed = sample mean; Red = 95% CI; Blue = null value"
  )
```

**Prompt (2–3 sentences):** Does the plot agree with your p-value-based conclusion? Explain.

> *Yes my plot agrees with my conclusion. My plot shows that my predicted value is far higher than the mean, and sits well outside of the 95% confidence interval. The likelihood of the true value being this high is essentially 0.*

---

## 10) Try a second claim (short extension)

Keep the same variable, but test a **different null value**.

```{r second_claim, echo=TRUE, eval=FALSE}
# GOAL: Test a second claim using the same sample and bootstrap SE.

mu0_alt <- 6

z_stat_alt <- (sample_mean - mu0_alt) / boot_se
p_value_alt <- 2 * pnorm(-abs(z_stat_alt))

tibble(
  mu0_alt = mu0_alt,
  z_stat_alt = z_stat_alt,
  p_value_alt = p_value_alt
)
```

**Prompt (3–4 sentences):** How did changing the null value change the conclusion? What does this tell you about the relationship between the null hypothesis and the observed estimate?

> *Changing the null value vastly improved my p-value. This tells me that the relationship between the null hypothesis and the observed estimate is dependant on the predicted null hypothesis. The observed estimate will not change unless more samples are taken, or samples are removed. Thus, the null hypothesis, which is variable and dependent on who is doing the prediction, is the determining factor for their relationship.*

---

# Part II — Comparing group means with t-tests

In Part I, you tested a claim about **one population mean**.

Now you will extend the same logic to compare **two group means**.

The key idea is:

> A two-sample t-test asks whether the difference between two group means is large relative to the uncertainty in that difference.

You will first work with a built-in grouping variable (**day vs night**), then choose a second grouping variable of your own.

---

## 11) Create a day/night grouping variable

If you already created `day_night` earlier, you can reuse it. Otherwise, create it here.

```{r make_daynight, echo=TRUE, eval=FALSE}
# GOAL: create a day/night grouping variable from `daytime`.

df <- df %>%
  mutate(
    day_night = if_else(daytime == 1, "Day", "Night")
  ) %>% 
  filter(!is.na(day_night))


df %>% count(day_night)
```

---

## 12) Choose a response variable for the t-test

Pick one numeric response variable to compare between groups.

Possible examples:
- `ta`
- `swc`
- `fc`
- `gpp`
- `le`

```{r choose_response_ttest, echo=TRUE, eval=FALSE}
response_var <- "ta"

# CHECK: does this variable exist?
response_var %in% names(df)
```

---

## 13) Visualize the response by day/night

Create a plot that shows how your response variable differs between Day and Night.

You may use:
- `geom_boxplot()`
- `geom_violin()`
- `geom_jitter()`
- or a combination

```{r plot_daynight, echo=TRUE, eval=FALSE}
# GOAL: visualize the response by day/night group.

df %>%
  filter(!is.na(.data[[response_var]])) %>%
  ggplot(aes(x = day_night, y = ta, fill = ta)) +
  geom_boxplot(alpha = 0.8) +
  theme_classic(base_size = 18) +
  labs(
    x = NULL,
    y = response_var,
    title = "Response by day vs night"
  )
```

**Prompt (2–3 sentences):** Before running the test, what do you predict?  
Do the groups look different in center? In spread?

> *The response variable look fairly similar in the center. Although their averages do not line up exactly, they have a lot of overlap in their spread. Because of this I predict that they are not statistically different.*

---

## 14) Summarize the two groups

Compute the group means, SDs, and sample sizes.

```{r summarize_daynight, echo=TRUE, eval=FALSE}
# GOAL: summarize the response variable by day/night group.

group_summary <- df %>%
  filter(!is.na(.data[[response_var]])) %>%
  group_by(day_night) %>%
  summarise(
    mean_response = mean(.data[[response_var]]),
    sd_response = sd(.data[[response_var]]),
    n = n(),
    .groups = "drop"
  )

group_summary
```

---

## 15) State the hypotheses for the day/night comparison

Write hypotheses in words and symbols.

Use this structure:

- Null: the two group means are equal
- Alternative: the two group means are different

Fill in:

> My response variable is `air temperature`

> $H_0$: `The average air temperature is the same at night and during the day`

> $H_A$: `The average air temperature is different during the day and at night`

---

## 16) Run a two-sample t-test

Use `t.test()` to compare the means of your response variable between Day and Night.

```{r ttest_daynight, echo=TRUE, eval=FALSE}
# GOAL: run a two-sample t-test by formula.

t.test(
  ta ~ day_night,
  data = df
)
```

**Prompt (4–6 sentences):** Report the p-value and confidence interval.  
Would you reject the null hypothesis at $\alpha = 0.05$?  
What does the result imply about day vs night differences in your chosen variable?

> *The p-value is < 2.2e-16, and the confidence interval is 5.573716 : 6.184308. I would reject the null hypothesis, due to the fact that the p-value is less than 0.05 and that the confidence interval does not include 0. This implies that the mean air temperature during the day in 2023 was different than the mean air temperature at night during 2023.*

---

## 17) Connect the t-test to a confidence interval

Extract the confidence interval and interpret it as a difference in means.

```{r ttest_store, echo=TRUE, eval=FALSE}
# GOAL: store the t-test result in an object.

tt_daynight <- t.test(
  ta ~ day_night,
  data = df
)

tt_daynight$conf.int
```

**Prompt (2–3 sentences):** Does the confidence interval include 0?  
How does that connect to the hypothesis test decision?

> *My confidence interval does not include 0. If the confidence interval included 0 then it would mean that there is a possibility that there is 0 difference between the means of the group, meaning that we would not be able to reject the null hypothesis. Because it does not include 0, it indicates that there is a difference between the means of the two groups.*

---

## 18) Bootstrap view of the same comparison (conceptual extension)

Create bootstrap distributions of the **group means** for Day and Night separately.

```{r bootstrap_group_means, echo=TRUE, eval=FALSE}
set.seed(9999)

B <- 3759   # e.g., 2000

x_day <- df %>%
  filter(day_night == "Day") %>%
  pull(ta) %>%
  na.omit()

x_night <- df %>%
  filter(day_night == "Night") %>%
  pull(ta) %>%
  na.omit()

boot_day <- replicate(
  B,
  mean(sample(x_day, size = length(x_day), replace = TRUE))
)

boot_night <- replicate(
  B,
  mean(sample(x_night, size = length(x_night), replace = TRUE))
)

boot_groups <- bind_rows(
  tibble(group = "Day", mean_est = boot_day),
  tibble(group = "Night", mean_est = boot_night)
)
```

```{r plot_boot_groups, echo=TRUE, eval=FALSE}
# GOAL: visualize the bootstrap distributions by group.

ggplot(boot_groups, aes(x = mean_est, fill = group)) +
  geom_histogram(alpha = 0.45, bins = 50, position = "identity") +
  theme_classic(base_size = 18) +
  labs(
    x = paste0("Bootstrap mean of ", response_var),
    y = "Count",
    title = "Bootstrap distributions of group means"
  )
```

**Prompt (3–4 sentences):** How does this bootstrap visualization help you understand the t-test result?  
Do the plausible group means strongly overlap, or are they clearly separated?

> *My t.test results indicated that the means of my two groups are not equivalent, rejecting the null hypothesis. This visualization shows me two groups that have no overlap and centers that are separated by several degrees. This graph visually shows how the means differ by large margins.*

---

## 19) Choose a second grouping variable of your own

Now define **one additional grouping variable** and repeat the t-test workflow.

Possible ideas:
- month grouped into two seasons (e.g., winter vs summer)
- a high/low split of another continuous variable
- a wet/dry split of SWC
- a daytime subset grouped by month category

You must:
- define the grouping variable clearly
- explain why the grouping is scientifically meaningful

```{r choose_second_group, echo=TRUE, eval=FALSE}
# GOAL: create your own two-level grouping variable.

df <- df %>%
  mutate(
    my_group = if_else(month %in% c("Nov", "Dec", "Jan", "Feb", "Mar", "Apr"), "Cold", "Warm"
  ))

view(df)

df %>% count(my_group)
```

**Prompt (2–3 sentences):** Why did you choose this grouping variable? What scientific contrast are you testing?

> *I chose this variable to compare the coldest six months of the year with the warmest six months of the year. This is a test to determine whether the average air temperature in the coldest six months of the year 2023 is equivalent to the average air temperature in the warmest six months of the year 2023.*

---

## 20) Visualize and test your second grouping variable

Repeat the workflow:
1. visualize
2. summarize
3. state hypotheses
4. run `t.test()`
5. interpret

```{r visualize_second_group, echo=TRUE, eval=FALSE}
# GOAL: visualize your response by your chosen grouping variable.

ggplot(df, aes(x = my_group, y = ta, fill = my_group)) +
  geom_boxplot(alpha = 0.8) +
  theme_minimal() +
  labs(
    x = NULL,
    y = response_var
  )
```

```{r summarize_second_group, echo=TRUE, eval=FALSE}
# GOAL: summarize your response variable by your chosen grouping variable.

group_summary <- df %>%
  filter(!is.na(.data[[response_var]])) %>%
  group_by(my_group) %>%
  summarise(
    mean_response = mean(.data[[response_var]]),
    sd_response = sd(.data[[response_var]]),
    n = n(),
    .groups = "drop"
  )

group_summary
```

```{r ttest_second_group, echo=TRUE, eval=FALSE}
# GOAL: run a two-sample t-test for your chosen grouping variable.

t.test(
  ta ~ my_group,
  data = df
)
```

**Prompt (5–6 sentences):** Compare the day/night t-test to your second grouping-variable test.  
Which contrast was stronger? Which was more scientifically meaningful?  
Did the two tests produce similar or different conclusions?

> *The two tests resulted in the same level of significance, indicating similar contrast. Both results produced a p-value of < 2.2e-16, capping out R's p-value. Because of this I can say with high confidence that the mean are temperature of the warm and cold months are not equivalent, and the mean air temperature of night and day are not equivalent. When comparing the two tests I cannot say whether there is a greater contrast in means in one test over the other, due to the high level of contrast found in both tests. I can say that both comparisons had a high level of contrast between the two groups*

---

## 21) Final reflection

Answer in **4–6 sentences**:

- How is a two-sample t-test similar to the one-mean hypothesis test from Part I?
- What changes when we move from “testing one mean” to “testing a difference in means”?
- How did the bootstrap group-mean plot help your understanding?

> *The t-test is a test the null hypothesis that two means are equal, with rejecting the null being the two samples means differ from one another. The one mean test tests the null hypothesis that the true mean of a sample is equal to a predicted value, with rejecting the null stating that the true mean is not equal to that value. They are similar in that both tests are with respect to the mean of the group or groups being tested. Additionally they compare two values and determine if they are equal. In the case of the one mean test we are comparing the true mean to the predicted mean, and in the case of the t.test we are comparing the means of the two sample groups. When we move from the one mean test to the t.test we change what we are comparing, where the t.test are saying that the means of two groups are equal or not, in the one mean test we are saying that the mean of one group is equal or not to a predicted value. *

---

# Submission

- Knit your `.Rmd` to HTML.
- Commit and push both the `.Rmd` and knitted `.html` to your GitHub repo.
- Submit the GitHub repository link on Canvas.
